Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-12 concurrency/CONCUR 2023/paper.tex
transcribed Lennart's comments and added some of my own

%TODO mandatory: add short abstract of the document
\begin{abstract}
... 
\end{abstract}

\section{Introduction}
\label{sec:introduction}
Krishna et al. proposed concurrent search structure templates~\cite{templates} as a method for separating the proof of correctness of a concurrent access method (optimistic concurrency, hand-over-hand locking, and internal links) from the proof of correctness of the underlying data structure (linked list, hashtable, and B-tree). Ideally, it should be possible to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns and obtain $n \times m$ verified concurrent data structures. However, in practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.

\lb{say something}

@@

\section{Background}
\subsection{Concurrent Search Structure Templates}
A search structure is a data structure designed to efficiently store and retrieve data based on specific search criteria. These structures organize data to enable quick and efficient search, insertion, deletion, and traversal operations. Concurrent search structures, on the other hand, are data structures accessible and modifiable by multiple threads or processes concurrently. However, designing concurrent search structures presents significant challenges, including ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory. Additionally, these structures must ensure compatibility with diverse hardware and software platforms.

@@

\section{Search Structure Templates}
The C implementation of a node in our template style is:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node_t {node *t; lock_t *lock;} node_t;
typedef struct pn {struct node_t *p; struct node_t *n;} pn;
\end{lstlisting}
\than{I changed \lstinline{tree_t} into \lstinline{node_t}, and \lstinline{tree} into \lstinline{node}, since concurrent search structure will not depend on particular data structures.}

@@

Figure \ref{traverse_lock} presents two versions of the \lstinline{traverse} function: one written in C (Figure \ref{traverse_lock_a}) and the other in an ML-like language (Figure \ref{traverse_lock_b}). The algorithm assumes that certain helper functions must be provided by the implementation that satisfy certain criteria. One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$ belonging to the set of key values.

@@

\begin{figure}[ht]
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		\label{traverse_lock_a}
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in ML-like language \cite{krishna2019compositional}}
		\label{traverse_lock_b}	
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms}
	\label{traverse_lock}
\end{figure}

To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}. The \lstinline{pn} struct includes two fields, namely \lstinline{p} and \lstinline{n}, which respectively represent the current and next node. Additionally, we require a target key \lstinline{k} and a dummy node value \lstinline{value} (referred to as 'dummy' because it is not actually used in \lstinline{traverse}). We include this dummy node to maintain consistency with other implementations. Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findNext}. The purpose of \lstinline{findNext} is to determine the next node to be visited during the traversal. In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}. Alternatively, it terminates if \lstinline{findNext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values. The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal. If it fails to do so, it returns false, which other functions can use to make decisions.

@@

\subsection{Give-Up}
The give-up template is an algorithm for implementing concurrent search in a tree data structure. Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires locks when loading or storing to a node. This reduces the amount of time that locks are held and makes the algorithm more scalable in highly-contended scenarios. Each node has a range field that serves as a reference point for threads to determine if they are still on the correct path. If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again. This behavior aligns with the give-up template algorithm's approach, in which threads terminate their search upon encountering obstacles or deviating from the expected path.

\begin{figure}[!ht]
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
		\label{traverse_giveup_a}
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in ML-like language}
		\label{traverse_giveup_b}	
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the give-up template algorithms}
	\label{traverse_giveup}
\end{figure}

Figure \ref{traverse_giveup_b} depicts the give-up template algorithm, which is presented in an ML-like language and described in \cite{krishna2019compositional}. The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section. The purpose of the \lstinline{inRange} function is to determine whether the key value $\texttt{k}$ falls within the range of node $\texttt{n}$. If the key value is outside the node's range, the search is terminated, and the thread is restarted.
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {node *t; lock_t *lock; int min; int max;} node_t;
\end{lstlisting}

@@

>>>>

%TODO mandatory: add short abstract of the document
\begin{abstract}
Concurrent search structure templates are a technique for separating the verification of a concurrent data structure into concurrency-control and data-structure components, which can then be modularly combined with no additional proof effort. In this paper, we implement the template approach in the Verified Software Toolchain, and use it to prove correctness of C implementations of concurrent data structures. %say more
\end{abstract}

\section{Introduction}
\label{sec:introduction}
Krishna et al. proposed concurrent search structure templates~\cite{templates} as a method for separating the proof of correctness of a concurrent access method (e.g. optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of correctness of the underlying data structure (e.g. linked list, hashtable, B-tree). The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency. In theory, this could allow us to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns, and immediately obtain $n \times m$ verified concurrent data structures. In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model.

@@

\section{Background}
\subsection{Concurrent Search Structure Templates}
\wm{This is probably more background than we need on search structures.}
A search structure is a data structure designed to efficiently store and retrieve data based on specific search criteria. These structures organize data to enable quick and efficient search, insertion, deletion, and traversal operations. Concurrent search structures, on the other hand, are data structures accessible and modifiable by multiple threads or processes concurrently. However, designing concurrent search structures presents significant challenges, including ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory, including cache behavior. \wm{cite Masstree?} Additionally, these structures must ensure compatibility with diverse hardware and software platforms.

@@

\section{Search Structure Templates in VST}
The C implementation of a node in our template style is:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node_t {node *t; lock_t *lock;} node_t;
\end{lstlisting}
\than{I changed \lstinline{tree_t} into \lstinline{node_t}, and \lstinline{tree} into \lstinline{node}, since concurrent search structure will not depend on particular data structures.}

@@

Figure \ref{traverse_lock} shows the \lstinline{traverse} function as presented by Krishna et al. (Figure \ref{traverse_lock_a}) and the corresponding C implementation (Figure \ref{traverse_lock_b}). The C implementation uses a struct
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct pn {struct node_t *p; struct node_t *n;} pn;
\end{lstlisting}
to mimic the pair of nodes \lstinline{(p, n)} returned by the functional implementation, where \lstinline{n} is the current node and \lstinline{p} is its parent. The algorithm assumes that certain helper functions must be provided by the implementation that satisfy certain criteria. One such function, namely \lstinline{findNext}, is utilized by the algorithm to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$. In the functional version, \lstinline{findNext} returns an \lstinline{option node}; in C, it instead returns a boolean and, if a next node is found, modifies \lstinline{pn->n}.

@@

\begin{figure}[ht]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional} \wm{Shouldn't it be lock and then unlock?}}
		\label{traverse_lock_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		\label{traverse_lock_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms}
	\label{traverse_lock}
\end{figure}

To implement the template methodology for concurrent data structure implementations in C, we have implemented the \lstinline{traverse} function in C, as depicted in Figure \ref{traverse_lock_a}. Additionally, we require a target key \lstinline{k} and a dummy node value \lstinline{value} (referred to as 'dummy' because it is not actually used in \lstinline{traverse}). We include this dummy value to maintain consistency with other implementations \wm{What other implementations?}. Our version of the \lstinline{traverse} function closely resembles the one written in an ML-like language, as it relies on a helper function called \lstinline{findNext}. The purpose of \lstinline{findNext} is to determine the next node to be visited during the traversal. In our implementation, the \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL}. Alternatively, it terminates if \lstinline{findNext} fails to find a next node that includes the key value $\texttt{k}$ in the set of key values. The \lstinline{traverse} function returns a Boolean value indicating whether it has successfully found the next node during traversal. If it fails to do so, it returns false, which other functions can use to make decisions.

@@

\subsection{Give-Up}
The give-up template is an algorithm for implementing concurrent search in a tree data structure. Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template takes an \emph{optimistic concurrency control} approach \lb{c.f. RCU}, only acquiring locks when loading or storing to a node. This reduces the amount of time that locks are held and makes the algorithm more scalable in highly-contended scenarios. Each node has a range field that serves as a reference point for threads to determine if they are still on the correct path. If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again. This behavior aligns with the give-up template algorithm's approach, in which threads terminate their search upon encountering obstacles or deviating from the expected path.

\begin{figure}[!ht]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in ML-like language}
		\label{traverse_giveup_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
		\label{traverse_giveup_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the give-up template algorithms}
	\label{traverse_giveup}
\end{figure}

Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented. The give-up template algorithm uses a helper function called \lstinline{inRange}, in addition to the \lstinline{findNext} function discussed earlier in the lock-coupling section. The purpose of the \lstinline{inRange} function is to determine whether the key value $\texttt{k}$ falls within the range of node $\texttt{n}$. If the key value is outside the node's range, the search is terminated, and the thread is restarted.

\wm{Describe why we need extra pieces in the struct, and how this relates to the other version.}
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {node *t; lock_t *lock; int min; int max;} node_t;
\end{lstlisting}

@@

\wm{I think this is the wrong organization. The proofs of traverse should be presented in the previous section, where we describe traverse. This section should show findNext and insert/lookup for specific data structure(s), and then describe how we combine them into top-level specs.}
